Neutral models of \textit{de novo} gene emergence show that gene evolution has a preferred trajectory

### 2022-11-10 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

If a gene is found in one species but not in sister taxa, then the most common assumption is that this gene emerged only in one specific species. We asked what is the chance that the gene was present in a common ancestor, but was lost in all lineages except one. To this end we calculated the probability of gene emergence and gene loss (Materials and Methods). Specifically, the probability of gene gain can be defined as the sum of three probabilities. First is the probability that an ORF already exists, and a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{eqrnagain}) but doesn't disrupt the ORF ($P_\textit{ORF-stay}$, \autoref{eqorfstay}). Second is the probability that the DNA is transcribed, and an ORF emerges due to mutations ($P_\textit{ORF-gain}$, \autoref{eqorfgain}) while transcription stays intact ($P_\textit{RNA-stay}$, \autoref{eqrnastay}). Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small, and negligible). The following equation describes the probability of gene gain:

@@

For a \textit{de novo} gene to emerge from a non-genic DNA sequence, both transcription and ORF need to emerge. That is, probability of gene emergence is equal to the product of probabilities of transcription gain and ORF gain. Thus it may appear that the order of the occurrence of these two events does not matter. However, gene emergence is much more likely when one of the two features already exist (\autoref{genegaineq}) and therefore it has two possible trajectories -- ORF emerges first (ORF-first) or transcription emerges first (RNA-first). To understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated two probabilities. First is the probability that ORF exists and  mutations cause gain of transcription but no disruption of the ORF. This probability denotes the trajectory where ORF emerges first ($P_\textit{ORF-first}$; \autoref{orffirsteq1}).

@@

The second probability ($P_\textit{RNA-first}$; \autoref{rnafirsteq1}) that denotes the trajectory where transcription emerges first, is the probability that DNA is already transcribed and mutations cause a gain of ORF but do not eliminate transcription.

@@

\begin{table}[H]
\centering
\begin{tabular}{c c}
\toprule
Position  & Accepted nucleotides \\ \midrule
1         & T$\gg$C$>$A,G        \\ \midrule
2         & A$\gg$T               \\ \midrule
3         & T$\gg$A,C             \\ \midrule
4         & A$\gg$T               \\ \midrule
5         & T/A                   \\ \midrule
6         & A$\gg$G$>$C/T         \\ \midrule
7         & A/T$>$G$>$C           \\ \midrule
8         & G/A$>$C/T             \\ \bottomrule
\end{tabular}
\caption{Nucleotide residues that are tolerated at different positions in a TATA-box \citep[based on][]{tata1}}
\label{tatatab}
\end{table}

>>>>

If a gene is found in one species but not in sister taxa, then the most common assumption is that this gene emerged only in one specific species. We asked what is the chance that the gene was present in a common ancestor, but was lost in all lineages except one. To this end we calculated the probability of gene emergence and gene loss (Materials and Methods). Specifically, the probability of gene gain can be defined as the sum of three probabilities. First is the probability that an ORF already exists, and a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{eqrnagain}) but doesn't disrupt the ORF ($P_\textit{ORF-stay}$, \autoref{eqorfstay}). Second is the probability that the DNA is transcribed, and an ORF emerges due to mutations ($P_\textit{ORF-gain}$, \autoref{eqorfgain}) while transcription stays intact ($P_\textit{RNA-stay}$, \autoref{eqrnastay}). Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small and is negligible). The following equation describes the probability of gene gain:

@@

For a \textit{de novo} gene to emerge from a non-genic DNA sequence, both transcription and ORF need to emerge. That is, probability of gene emergence is equal to the product of probabilities of transcription gain and ORF gain. Thus it may appear that the order of the occurrence of these two events does not matter. However, gene emergence is much more likely when one of the two features already exist (\autoref{genegaineq}) and therefore it has two possible trajectories -- ORF emerges first (ORF-first) or transcription emerges first (RNA-first). To understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated two probabilities. First is the probability that ORF exists and  mutations cause gain of transcription but no disruption of the ORF. This probability denotes the trajectory where ORF emerges first ($P_\textit{ORF-first}$, \autoref{orffirsteq1}).

@@

The second probability ($P_\textit{RNA-first}$, \autoref{rnafirsteq1}) that denotes the trajectory where transcription emerges first, is the probability that DNA is already transcribed and mutations cause a gain of ORF but do not eliminate transcription.

@@

\begin{table}[!t]
\centering
\begin{tabular}{c c}
\toprule
Position  & Accepted nucleotides \\ \midrule
1         & T$\gg$C$>$A,G        \\ \midrule
2         & A$\gg$T               \\ \midrule
3         & T$\gg$A,C             \\ \midrule
4         & A$\gg$T               \\ \midrule
5         & T/A                   \\ \midrule
6         & A$\gg$G$>$C/T         \\ \midrule
7         & A/T$>$G$>$C           \\ \midrule
8         & G/A$>$C/T             \\ \bottomrule
\end{tabular}
\caption{Nucleotide residues that are tolerated at different positions in a TATA-box \citep[based on][]{tata1}}
\label{tatatab}
\end{table}
